---
title: "Whitman & Werdel (2024) code"
author: "Daniel Whitman"
date: "2024-04-12"
output: html_document
---

# 1. Load packages and data.

```{r}
library(ggplot2)
library(dplyr)
library(AICcmodavg)
library(broom)

transects <- read.csv("all_transects.csv")

# Rename columns and remove unnecessary columns.
transects <- rename(transects, min.T= min.T.degrees.C) 
transects <- rename(transects, max.T= max.T.degrees.C)
transects <- rename(transects, min.T.est = estimated.min.T.degrees.C)
transects <- rename(transects, squir.dens = squir.dens.sq.km)
transects <- rename(transects, tree.dens = tree.density.sq.km)
transects <- rename(transects, oak.dens = oak.density.sq.km)
transects <- rename(transects, length=length.km)
transects <- rename(transects, area = area.sq.km)

transects <- transects %>%
  select(-min.T.degrees.F) %>%
  select(-estimated.min.T.degrees.F) %>%
  select(-max.T.degrees.F)

transects <- transects %>%
  mutate(prop.oak = num.oaks/num.trees)
```

# 2. Calculate Pearson's Correlation Coefficients to check for colinearity

```{r}
# Create a dataframe in which transects without recorded temperature min/max 
# have been removed to calculate PCCs for (non-estimated) min/max temperature.
transects_no_na <- subset(transects, !is.na(min.T))
transects_no_na <- subset(transects_no_na, !is.na(max.T))

cc_min_max <- cor(transects_no_na$min.T, transects_no_na$max.T)
cc_min_hr <- cor(transects_no_na$min.T, transects_no_na$hour)
cc_max_hr <- cor(transects_no_na$max.T, transects_no_na$hour)

cc_hr_dur <- cor(transects$hour, transects$dur.min)
cc_hr_estT <- cor(transects$hour, transects$min.T.est)
cc_len_a <- cor(transects$length, transects$area)
cc_tree_oak <- cor(transects$tree.dens, transects$oak.dens)
cc_oak_prop <- cor(transects$oak.dens, transects$prop.oak)
cc_tree_a <- cor(transects$tree, transects$area)
cc_tree_len <- cor(transects$tree, transects$length)
cc_a_dur <- cor(transects$area, transects$dur.min)
cc_len_dur <- cor(transects$length, transects$dur.min)
cc_prop_a <- cor(transects$prop.oak, transects$area)

PCC <- c(cc_hr_dur, cc_hr_estT, cc_min_max, cc_min_hr, cc_max_hr, cc_len_a, cc_tree_oak, cc_oak_prop, cc_tree_a, cc_tree_len, cc_a_dur, cc_len_dur, cc_prop_a)
parameter.1 <- c("hour", "hour", "min.T", "min.T", "max.T", "length", "tree.dens", "oak.dens", "tree.dens", "tree.dens", "area", "length", "prop.oak")
parameter.2 <- c("dur.min", "min.T.est", "max.T", "hour", "hour", "area", "oak.dens", "prop.oak", "area", "length", "dur.min", "dur.min", "area")
correlations <- data.frame(parameter.1, parameter.2, PCC)
print(correlations)
```

# 3. Initial round of model construction

```{r}
# Check effect of categorical time of day
anova_result <- aov(squir.dens ~ time.of.day, data = transects)
summary(anova_result)
TukeyHSD(anova_result, conf.level=.95)
```

```{r}
# Model squirrels by time and temperature.
## Time
model_1a <- lm(squir.dens ~ time.of.day + length, data=transects)
model_1b <- lm(squir.dens ~ hour + length, data=transects)
model_1c <- lm(squir.dens ~ min.T + length, data=transects)
model_1d <- lm(squir.dens ~ max.T + length, data=transects)
model_1e <- lm(squir.dens ~ min.T.est + length, data=transects)

model_2a <- lm(squir.dens ~ time.of.day + area, data=transects)
model_2b <- lm(squir.dens ~ hour + area, data=transects)
model_2c <- lm(squir.dens ~ min.T + area, data=transects)
model_2d <- lm(squir.dens ~ max.T + area, data=transects)
model_2e <- lm(squir.dens ~ min.T.est + area, data=transects)

model_3a <- lm(squir.dens ~ tree.dens + prop.oak, data=transects)
model_3b <- lm(squir.dens ~ oak.dens + prop.oak, data=transects)

model_4a <- lm(squir.dens ~ tree.dens*prop.oak, data=transects)
model_4b <- lm(squir.dens ~ oak.dens*prop.oak, data=transects)

model_5a <- lm(squir.dens ~ tree.dens + time.of.day, data=transects)
model_5b <- lm(squir.dens ~ tree.dens + hour, data=transects)
model_5c <- lm(squir.dens ~ tree.dens + min.T, data=transects)
model_5d <- lm(squir.dens ~ tree.dens + max.T, data=transects)
model_5e <- lm(squir.dens ~ tree.dens + min.T.est, data=transects)

model_6a <- lm(squir.dens ~ oak.dens + time.of.day, data=transects)
model_6b <- lm(squir.dens ~ oak.dens + hour, data=transects)
model_6c <- lm(squir.dens ~ oak.dens + min.T, data=transects)
model_6d <- lm(squir.dens ~ oak.dens + max.T, data=transects)
model_6e <- lm(squir.dens ~ oak.dens + min.T.est, data=transects)

model_7a <- lm(squir.dens ~ tree.dens + dur.min, data=transects)
model_7b <- lm(squir.dens ~ oak.dens + dur.min, data=transects)

model_8a <- lm(squir.dens ~ tree.dens*time.of.day, data=transects)
model_8b <- lm(squir.dens ~ tree.dens*hour, data=transects)
model_8c <- lm(squir.dens ~ tree.dens*min.T, data=transects)
model_8d <- lm(squir.dens ~ tree.dens*max.T, data=transects)
model_8e <- lm(squir.dens ~ tree.dens*min.T.est, data=transects)

model_9a <- lm(squir.dens ~ oak.dens*time.of.day, data=transects)
model_9b <- lm(squir.dens ~ oak.dens*hour, data=transects)
model_9c <- lm(squir.dens ~ oak.dens*min.T, data=transects)
model_9d <- lm(squir.dens ~ oak.dens*max.T, data=transects)
model_9e <- lm(squir.dens ~ oak.dens*min.T.est, data=transects)

model_10a <- lm(squir.dens ~ tree.dens + poly(hour, 2), data=transects)
model_10b <- lm(squir.dens ~ tree.dens + poly(min.T.est, 2), data=transects)

model_11a <- lm(squir.dens ~ oak.dens + poly(hour, 2), data=transects)
model_11b <- lm(squir.dens ~ oak.dens + poly(min.T.est, 2), data=transects)

# single-variable models
m_min.T <- lm(squir.dens ~ min.T, data=transects)
m_max.T <- lm(squir.dens ~ max.T, data=transects)
m_min.T.est <- lm(squir.dens ~ min.T.est, data=transects)
m_hour <- lm(squir.dens ~ hour, data=transects)
m_dur.min <- lm(squir.dens ~ dur.min, data=transects)
m_length <- lm(squir.dens ~ length, data=transects)
m_area <- lm(squir.dens ~ area, data=transects)
m_tree.dens <- lm(squir.dens ~ tree.dens, data=transects)
m_oak.dens <- lm(squir.dens ~ oak.dens, data=transects)
m_prop.oak <- lm(squir.dens ~ prop.oak, data=transects)

# null and global models
null <- lm(squir.dens ~ 1, data = transects)
global <- lm(squir.dens ~ date + hour + time.of.day + dur.min + length + area + min.T.est + tree.dens + oak.dens + prop.oak, data=transects)
```

## Comparing models by AICc

```{r}
models <- list(model_1a, model_1b, model_1c, model_1d, model_1e, model_2a, model_2b, model_2c, model_2d, model_2e, model_3a, model_3b, model_4a, model_4b, model_5a, model_5b, model_5c, model_5d, model_5e, model_6a, model_6b, model_6c, model_6d, model_6e, model_7a, model_7b, model_8a, model_8b, model_8c, model_8d, model_8e, model_9a, model_9b, model_9c, model_9d, model_9e, model_10a, model_10b, model_11a, model_11b, m_min.T, m_max.T, m_min.T.est, m_hour, m_dur.min, m_length, m_area, m_tree.dens, m_oak.dens, m_prop.oak, null, global)

mod_names <- c('model_1a', 'model_1b', 'model_1c', 'model_1d', 'model_1e', 'model_2a', 'model_2b', 'model_2c', 'model_2d', 'model_2e', 'model_3a', 'model_3b', 'model_4a', 'model_4b', 'model_5a', 'model_5b', 'model_5c', 'model_5d', 'model_5e', 'model_6a', 'model_6b', 'model_6c', 'model_6d', 'model_6e', 'model_7a', 'model_7b', 'model_8a', 'model_8b', 'model_8c', 'model_8d', 'model_8e', 'model_9a', 'model_9b', 'model_9c', 'model_9d', 'model_9e', 'model_10a', 'model_10b', 'model_11a', 'model_11b', 'm_min.T', 'm_max.T', 'm_min.T.est', 'm_hour', 'm_dur.min', 'm_length', 'm_area', 'm_tree.dens', 'm_oak.dens', 'm_prop.oak', 'null', 'global')

AICc_results_1 <- aictab(cand.set = models, modnames = mod_names)
print(AICc_results_1)

# Models incorporating minimum or maximum temperature have lower AICc than similar models using hour. However, using estimated mimumum temperature instead, similar models using hour have lower AICc. 
```

## 2nd round of model constrution and evaluation

```{r}
model_12a <- lm(squir.dens ~ tree.dens + hour + length, data=transects)
model_12b <- lm(squir.dens ~ oak.dens + hour + length, data=transects)
model_12c <- lm(squir.dens ~ tree.dens + hour + prop.oak, data=transects)
model_12d <- lm(squir.dens ~ oak.dens + hour + prop.oak, data=transects)

model_13a <- lm(squir.dens ~ tree.dens + poly(hour, 2) + length, data=transects)
model_13b <- lm(squir.dens ~ oak.dens + poly(hour, 2) + length, data=transects)
model_13c <- lm(squir.dens ~ tree.dens + poly(hour, 2) + prop.oak, data=transects)
model_13d <- lm(squir.dens ~ oak.dens + poly(hour, 2) + prop.oak, data=transects)

model_14a <- lm(squir.dens ~ tree.dens + hour + length + prop.oak, data=transects)
model_14b <- lm(squir.dens ~ oak.dens + hour + length + prop.oak, data=transects)

model_15a <- lm(squir.dens ~ tree.dens*prop.oak + length, data=transects)
model_15b <- lm(squir.dens ~ oak.dens*prop.oak + length, data=transects)
model_15c <- lm(squir.dens ~ tree.dens*prop.oak + hour, data=transects)
model_15d <- lm(squir.dens ~ oak.dens*prop.oak + hour, data=transects)

model_16a <- lm(squir.dens ~ tree.dens*hour + length, data=transects)
model_16b <- lm(squir.dens ~ oak.dens*hour + length, data=transects)
model_16c <- lm(squir.dens ~ tree.dens*hour + dur.min, data=transects)
model_16d <- lm(squir.dens ~ oak.dens*hour + dur.min, data=transects)

model_17a <- lm(squir.dens ~ tree.dens*prop.oak + hour + length, data=transects)
model_17b <- lm(squir.dens ~ oak.dens*prop.oak + hour + length, data=transects)
```


## Compare models by AICc (including models from the first round of model building with lowest AICc values)

```{r}
models_2 <- list(model_12a, model_12b, model_12c, model_12d, model_13a, model_13b, model_13c, model_13d, model_14a, model_14b, model_15a, model_15b, model_15c, model_15d, model_16a, model_16b, model_16c, model_16d, model_17a, model_17b, model_8b, model_9b, model_5b, model_2b, model_1b, m_min.T.est, m_hour, m_dur.min, m_length, m_area, m_tree.dens, m_oak.dens, m_prop.oak, null, global)

mod_names_2 <- c('model_12a', 'model_12b', 'model_12c', 'model_12d', 'model_13a', 'model_13b', 'model_13c', 'model_13d', 'model_14a', 'model_14b', 'model_15a', 'model_15b', 'model_15c', 'model_15d', 'model_16a', 'model_16b', 'model_16c', 'model_16d', 'model_17a', 'model_17b', 'model_8b', 'model_9b', 'model_5b', 'model_2b', 'model_1b', 'm_min.T.est', 'm_hour', 'm_dur.min', 'm_length', 'm_area', 'm_tree.dens', 'm_oak.dens', 'm_prop.oak', 'null', 'global')

AICc_results_2 <- aictab(cand.set = models_2, modnames = mod_names_2)
print(AICc_results_2)
```

## Comparing just the 5 highest-ranked models against single-variable, null, and global

```{r}
models_3 <- list(model_12a, model_16a, model_16b, model_17b, model_8b, m_min.T.est, m_hour, m_dur.min, m_length, m_area, m_tree.dens, m_oak.dens, m_prop.oak, null, global)

mod_names_3 <- c('model_12a', 'model_16a', 'model_16b', 'model_17b', 'model_8b', 'm_min.T.est', 'm_hour', 'm_dur.min', 'm_length', 'm_area', 'm_tree.dens', 'm_oak.dens', 'm_prop.oak', 'null', 'global')

AICc_results_3 <- aictab(cand.set = models_3, modnames = mod_names_3)
print(AICc_results_3)
```


# 4. Calculate Jacob's Selection Index of trees were occupied by a squirrel and that squirrels were nearest to.
```{r}
# Import data
all_trees <- read.csv('trees_by_species.csv')
occu_trees <- read.csv('squirrels_in_trees.csv')
near_trees <- read.csv('squirrels_near_trees.csv')
near_trees <- subset(near_trees, duplicate != "yes")

# Summarize number of occupied trees by species
occu_trees_1 <- occu_trees %>%
  group_by(sp_name) %>%
  summarize(Abundance = n_distinct(Date.Time))
occu_trees_2 <- occu_trees %>%
  group_by(Genus) %>%
  summarize(Abundance = n_distinct(Date.Time)) %>%
  rename(sp_name = Genus)
occu_trees_2 <- subset(occu_trees_2, sp_name != "Triadica sp.")
occu_trees_2 <- subset(occu_trees_2, sp_name != "Lagerstroemia sp.")
occu_trees_2 <- subset(occu_trees_2, sp_name != "Unknown")
occupied <- rbind(occu_trees_1, occu_trees_2)
sel_data <- all_trees %>% full_join(occupied, by="sp_name")
sel_data <- sel_data %>%
  rename(Available=Abundance.x) %>%
  rename(Used=Abundance.y) %>%
  mutate(Available=as.numeric(Available)) %>%
  mutate(Used=as.numeric(Used))
sel_data[is.na(sel_data)] <- 0

# Calculte Jacob's selection of trees occupied
## r = proportion of species out of occupied trees
## p = proportion of species out of available trees
## Jacob's: (r-p)/(r+p-2rp)
Jacob_Occupied <- numeric()
for(i in 1:nrow(sel_data)){
  r = sel_data[i, 3]/sum(sel_data$Used)
  p = sel_data[i, 2]/sum(sel_data$Available)
  J = (r - p)/(r + p - 2*r*p)
  Jacob_Occupied <- append(Jacob_Occupied, J)
}

sel_data <- cbind(sel_data, Jacob_Occupied)
## Just the tree species that had squirrels
short_sel <- subset(sel_data, Used>0)
```

```{r}
# Summarize nearest trees by species
near_trees_1 <- near_trees %>%
  group_by(sp_name) %>%
  summarize(Abundance = n_distinct(Date.Time))
near_trees_2 <- near_trees %>%
  group_by(Genus) %>%
  summarize(Abundance = n_distinct(Date.Time)) %>%
  rename(sp_name = Genus)
near_trees_1 <- subset(near_trees_1, sp_name != "")
near_trees_2 <- subset(near_trees_2, sp_name != "")
near <- rbind(near_trees_1, near_trees_2)

sel_data_2 <- all_trees %>% full_join(near, by="sp_name")
sel_data_2 <- sel_data_2 %>%
  rename(Available=Abundance.x) %>%
  rename(Used=Abundance.y) %>%
  mutate(Available=as.numeric(Available)) %>%
  mutate(Used=as.numeric(Used))
sel_data_2[is.na(sel_data_2)] <- 0

# Calculte Jacob's selection of trees nearest to squirrel
## Jacob's: (r-p)/(r+p-2rp)
Jacob_Near <- numeric()
for(i in 1:nrow(sel_data_2)){
  r = sel_data_2[i, 3]/sum(sel_data_2$Used)
  p = sel_data_2[i, 2]/sum(sel_data_2$Available)
  J = (r - p)/(r + p - 2*r*p)
  Jacob_Near <- append(Jacob_Near, J)
}

sel_data_2 <- cbind(sel_data_2, Jacob_Near)
## Just the species that actually had squirrels nearby
short_sel_2 <- subset(sel_data_2, Used>0) 
```

# Combine lists of indices for selection of nearest trees and occupied trees

```{r}
all_selections <- sel_data_2 %>% full_join(sel_data, by="sp_name")
all_selections <- all_selections %>%
  select(-Available.y) %>%
  rename(Available=Available.x) %>%
  rename(Occupied=Used.y) %>%
  rename(Near=Used.x)
## Just species that squirrels occupied or were near
short_selections <- short_sel_2 %>% full_join(short_sel, by="sp_name")
short_selections <- short_selections %>%
  select(-Available.y) %>%
  rename(Available=Available.x) %>%
  rename(Occupied=Used.y) %>%
  rename(Near=Used.x)
```
